---
title: "Policy Project"
output:
  html_document:
    theme: united
    highlight: 
    df_print: paged
    code_folding: "hide"
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project will explore the relationship between the increasing energy tarrifs and it's effect on the different types of expenditure from debit/credit cards.

Research question: How has the increasing price of energy (specifically for household energy) affected spending on different sectors?

The two datasets I will be using are:

* The CHAPS (Clearing House Automated Payment System ) dataset
* The ofgem retial energy prices data set 

# Background

The CHAPS data series are experimental faster indicators for monitoring UK spending on credit and debit cards. They track the daily CHAPS payments made by credit and debit card payment processors (often known as ‘merchant acquirers’) to around 100 major UK retail corporates. These payments are the proceeds of recent credit and debit card transactions made by customers at their stores (both physical and via telephone/online platforms). (taken from https://www.bankofengland.co.uk/payment-and-settlement/chaps-faster-indicator)

The ofgem retail energy prices dataset covers the Average standard variable tariff from Large legacy suppliers as well as the cheapeast tarrifs at the time. I will be using the standard average tarrifs from the large suppliers, as this will give me the best indication of the general populations energy tarrifs. 

Hopefully, by comparing the different types of expenditure over the last few years and comparing it to energy tarrif prices, I can identify whether there is a relationship between the two.
```{r, warning = FALSE, message = FALSE}
# load in packages needed
library("tidyverse")
library("ggplot2")
library("readxl")
library("openxlsx")

```

# Looking at CHAPS data

A look at the data after pre-processing: 

* 'staples' refers to companies that sell essential goods that households need to purchase, such as food and utilities
* 'work-related' refers to companies providing public transport or selling petrol
*  'delayable' refers to companies selling goods whose purchase could be delayed, such as clothing or furnishings
* 'social' refers to spending on travel and eating out

## Daily data, non-seasonally adjusted:

```{r, warning = FALSE, message = FALSE}
# load in data
excel_file <- "Data/ukspendingoncreditanddebitcardsdataset010923.xlsx"
data <- read_excel(excel_file, sheet = "Daily CHAPS indices")

# remove uneccessary info from headers
data <- data[3:nrow(data),]

# reset colnames
colnames(data)<- data[1,]
data<- data[-1,]
data$Date <- convertToDateTime(data$Date)

# change numbers from characters to numeric
numeric_columns <- c("Aggregate", "Delayable", "Social", "Staple", "Work Related")  # Replace with your column names
data[numeric_columns] <- lapply(data[numeric_columns], as.numeric)

# pivvot column to type column for ease of plotting
data <- pivot_longer(data, cols = numeric_columns, names_to = "Type")

head(data)
```

## Monthly data, non-seasonally adjusted:

```{r, warning = FALSE, message = FALSE}
monthly_data <- read_excel(excel_file, sheet = "Monthly CHAPS index")

# remove uneccessary info from headers
monthly_data <- monthly_data[3:nrow(data),]

# reset colnames
colnames(monthly_data)<- monthly_data[1,]
monthly_data<- monthly_data[-1,]
monthly_data$Month <- convertToDateTime(monthly_data$Month)

# change numbers from characters to numeric
numeric_columns <- c("Aggregate", "Delayable", "Social", "Staple", "Work Related")  # Replace with your column names
monthly_data[numeric_columns] <- lapply(monthly_data[numeric_columns], as.numeric)

# pivvot column to type column for ease of plotting
monthly_data <- pivot_longer(monthly_data, cols = numeric_columns, names_to = "Type")

head(monthly_data)

```

## Aggregate monthly data, seasonally adjusted:

```{r, warning = FALSE, message = FALSE}

monthly_data_sa <- read_excel(excel_file, sheet = "Monthly CHAPS index SA")

# remove uneccessary info from headers
monthly_data_sa <- monthly_data_sa[3:nrow(data),]

# reset colnames
colnames(monthly_data_sa)<- monthly_data_sa[1,]
monthly_data_sa<- monthly_data_sa[-1,]
monthly_data_sa$Month <- convertToDateTime(monthly_data_sa$Month)

# change numbers from characters to numeric

monthly_data_sa$Aggregate <- as.numeric(monthly_data_sa$Aggregate)

head(monthly_data_sa)

```

## Visualisations of the data: {.tabset} 

### Daily data 

```{r, warning = FALSE, message = FALSE}
ggplot(data, aes(x = Date, y = value)) +
  geom_line() +
  facet_wrap(~Type)+
  labs(y = "CHAPS index")
```

### Monthly data (non-sa)

```{r, warning = FALSE, message = FALSE}
ggplot(monthly_data, aes(x = Month, y = value)) +
  geom_line() +
  facet_wrap(~Type)+
  labs(y = "CHAPS index")
```

### Aggregate monthly data (sa)

```{r, warning = FALSE, message = FALSE}
ggplot(monthly_data_sa, aes(x = Month, y = Aggregate)) +
  geom_line() +
  labs(y = "CHAPS index") +
  labs(y = "Aggregate CHAPS index")
```


## Key insights

 * Looking at the plots for spending, it's clear that Christmas season has a strong effect on spending. Increasing aggregate, staple, delay able spending, and decreasing social/work related spending.
 * It is also clear that the pandemic had a clear impact on spending over the last few years, for example with the March 2020 lockdown Social spending decreased drastically, returning to regular levels      after the July 2021 lock down ended.
 
# Energy tarrif exploration

```{r, warning = FALSE, message = FALSE}
# load in the data
retail_data <- read.csv("Data/retail-price-comparison.csv")
retail_names <- c("Date", "Avg_st_large", "Avg_st_other", "Avg_fixed", "Cheap_tarr_large", "Cheap_tarr_all_supp", "Cheap_tarr_bask", "Default")
colnames(retail_data) <- retail_names
retail_data$Date <- format(as.POSIXct(retail_data$Date,format='%Y-%m-%d %H:%M:%S'),format='%Y-%m-%d')
retail_data$Date<- as.POSIXct(retail_data$Date, format="%Y-%m-%d", tz="UTC")

# filter for same dates as CHAPS data
retail_data <- retail_data %>% 
  filter(Date > "2020-01-12")

head(retail_data)
```

## Visualisations {.tabset}

### Average standard variable tarrif from large legacy companies

```{r, warning = FALSE, message = FALSE}
#simple plot
ggplot(retail_data, aes(x = Date,y = Avg_st_large)) +
  geom_line() + 
  labs(y = "Average standard variable tarrif (Pounds/year)")
```

### Average fixed tarrif

```{r, warning = FALSE, message = FALSE}
#simple plot
ggplot(retail_data, aes(x = Date,y = Avg_fixed)) +
  geom_line() +
  labs(y = "Average fixed tarrif (Pounds/year)")
```


### Tarrif cap 

```{r, warning = FALSE, message = FALSE}
#simple plot
ggplot(retail_data, aes(x = Date,y = Default)) +
  geom_line() +
  labs(y = "Tarrif cap (Pounds/year)")
```


## Key insights

* From the plots we can see that the average standard variable tarrifs nearly exactly matches those of the tarrif caps, with companies always increasing costs where possible.

# Looking at the trends together 

## Visualisations {.tabset}

### daily data vs std energy prices

```{r, warning = FALSE, message = FALSE} 
ggplot() +
  geom_line(data = data, aes(x = Date, y = value, color = "Spending")) +
  geom_line(data = retail_data, aes(x = Date, y = Avg_st_large/15, color = "Energy prices")) +
  scale_y_continuous(sec.axis = sec_axis(~.*15, name= "Average standard variable tarrif (Pounds/year))")) +
  facet_wrap(~Type) +
  labs(y = "CHAPS index") +
  theme(legend.position = c(0.9, 0.25))
```


### daily data vs fixed average

```{r, warning = FALSE, message = FALSE} 
ggplot() +
  geom_line(data = data, aes(x = Date, y = value, color = "Spending")) +
  geom_line(data = retail_data, aes(x = Date, y = Avg_fixed/15, color = "Energy prices")) +
  scale_y_continuous(sec.axis = sec_axis(~.*15, name= "Average fixed tarrif (Pounds/year)")) +
  facet_wrap(~Type)+
  labs(y = "CHAPS index") +
  theme(legend.position = c(0.9, 0.25))
```

### daily vs energy cap tarrif

```{r, warning = FALSE, message = FALSE} 
ggplot() +
  geom_line(data = data, aes(x = Date, y = value, color = "Spending")) +
  geom_line(data = retail_data, aes(x = Date, y = Default/15, color = "Energy prices")) +
  scale_y_continuous(sec.axis = sec_axis(~.*15, name= "Tarrif cap (Pounds/year)")) +
  facet_wrap(~Type)+
  labs(y = "CHAPS index") +
  theme(legend.position = c(0.9, 0.25))
```

### monthly sa aggregate data vs std energy prices

```{r, warning = FALSE, message = FALSE} 
ggplot() +
  geom_line(data = monthly_data_sa, aes(x = Month, y = Aggregate, color = "Spending")) +
  geom_line(data = retail_data, aes(x = Date, y = Avg_st_large/15, color = "Energy prices")) +
  scale_y_continuous(sec.axis = sec_axis(~.*15, name= "energy prices")) 
```

## Key insights

* Looking at the plots together, with the price increases often occurring around the Christmas period, it is hard to extract any meaningful insights from the non seasonally adjusted data. 
* With each increase in price there is often a rise in spending on staple products, but it's impossible to say with this data whether it is causative.
* The seasonally adjusted comparison plot also makes it hard to see any clear patterns, especially since the spending data is only available at aggregate level. 

# Limitations

CHAPS limitations:

* Only includes 100 major retailers - may create bias
* Only monitors credit/debit cards - excluding cash may bias towards higher socio-economic groups
* No seasonally adjusted data for the different types apart from aggregate, which makes it very hard to extract meaningful insights
* The groups are quite broad, which makes it hard to specifically target areas that may be effected by rising energy tarrifs.

Other:

* Data is also from the COVID-19 pandemic, which may have a real effect on how people spent money. E.g. less spending on hospitality/eating out due to being in lockdown. This effect may be making it harder to see how people's spending changes with varying energy tarrifs.
* With tight time constraints it was hard to find appropriate data within recent time-frames, as well as then carrying out detailed analysis on it.

# Conclusions, recommendations and future research

Looking at the non-seasonally adjusted data, it is especially hard to see how increasing prices are effecting spending, mainly due to the fact that increases often happen around Christmas which has a large confounding effect on spending.

In conclusion, looking at the aggregate seasonally adjusted data in comparison the average energy tarrif prices, it does not look like there is an easily identifiable correlation. 

In the future, with more time available, it would be great to explore this relationship between different socio-economic backgrounds. By looking at those from lower socio-economic backgrounds we may be able to identify how their spending changes, and identify potential areas which can be aided.

It would also be great to this with the categories further broken down, allowing us to see specifically which industries are affected the most by increasing energy prices.

# Sources

* Bank Of England CHAPS - https://www.bankofengland.co.uk/payment-and-settlement/chaps-faster-indicator
* Ofgem tarrif data (have to download chart as csv)- https://www.ofgem.gov.uk/energy-data-and-research/data-portal/all-available-charts?sort=created&page=7